---
title: "Class06_Homework_6002_yc1222"
author: "Youngmin Choi"
format: 
  html: 
    self-contained: true
---

```{r setup}
echo=TRUE
message=FALSE
warning= FALSE
library(tidyverse)     
library(RPostgres)     
library(connections)   
library(keyring)       
```

## I. Objective

-   Asthma is a chronic lung disease caused by inflammation around airways, making it hard to breathe, with symptoms such as coughing, wheezing, and chest tightness.
-   Asthma cannot be cured but it can be managed with the right treatment. (source: <https://www.who.int/news-room/fact-sheets/detail/asthma>)
-   Therefore, taking care of patients with asthma is important as this relates to the management or prevention of symptoms.
-   In this report, asthma patient data were analyzed by defining asthma patient population cohort and current care status of those patients, in order to evaluate plan for proper quality of care for asthma patients.

## II. Dataset

-   The dataset used for this analysis is the Synthea dataset, a synthetic electronic health records.

```{r package loading and database connection}

# Loading Synthea dataset
con <- connection_open(RPostgres::Postgres(),
          dbname = "",
          host = "",
          user = "",
          password = 

```

## III. Target Population Data Analysis

#### 1) Patient Filter Criteria

-   Patients with asthma diagnosis or screening.

    -   From conditions table/description column: containing 'Asthma' or 'asthma'
    -   From encounters table/reason description column: containing 'Asthma' or 'asthma'
    -   From procedures table/description column: containing 'Asthma' or 'asthma'
    -   From medications table/code: patients with history of asthma-related medications.

-   Following is the SQL code to include patients meeting at least one of the aforementioned standard, using the UNION operator. For the purpose of this analysis, only patients with data in the year 2022 were included.

-   A temporary table was created for analysis.

```{sql connection=con}
DROP TABLE IF EXISTS asthmaCohort
```

```{sql connection=con}
  WITH asthma_cohort AS (
  SELECT patient 
  FROM conditions 
  WHERE EXTRACT(YEAR FROM start) =2022
  AND (description LIKE '%Asthma%' OR description LIKE '%asthma%')

  UNION

  SELECT patient
  FROM encounters
  WHERE EXTRACT(YEAR FROM start) =2022
  AND (reasondescription LIKE '%Asthma%' OR reasondescription LIKE '%asthma%')

  UNION

  SELECT patient
  FROM procedures
  WHERE EXTRACT(YEAR FROM start) =2022
  AND (reasondescription LIKE '%Asthma%' OR reasondescription LIKE '%asthma%')
  
  UNION
  
  SELECT DISTINCT patient
  FROM medications
  WHERE code IN (95996, 349094, 616830, 1998774, 896209, 896006, 966529, 746815, 966524, 1547660, 1998772, 1536144, 630208, 859088, 745752, 351137, 351136, 245314, 2123111, 746763)
  )
 SELECT * INTO TEMP asthmaCohort FROM asthma_cohort
 
```

-   Confirm if the temporary table has been created successfully.

```{sql connection=con}
  SELECT * FROM asthmaCohort LIMIT 5
```

-   Confirmed if number of data (rows) matches the number of unique (distinct) patients in year 2022.\
-   Total number of patients in the asthma cohort is 2298.

```{sql connection=con}
  SELECT COUNT(*) AS rows, COUNT(DISTINCT patient) FROM asthmaCohort 
```

#### 2) Patient Demographics

**Overall Information**

-   Following is the average age, and gender ratio information of the asthma cohort patient group.

```{sql connection=con}
SELECT CAST(AVG(EXTRACT(year FROM AGE('2022-12-31', CAST(p.birthdate AS DATE)))) AS INT) AS average_age, 
(SUM(CASE WHEN p.gender='M' THEN 1 ELSE 0 END)*100)/(SELECT COUNT(*) FROM asthmaCohort) AS male_patient_ratio,
(SUM(CASE WHEN p.gender='F' THEN 1 ELSE 0 END)*100)/(SELECT COUNT(*) FROM asthmaCohort) AS female_patient_ratio 
FROM patients AS p
WHERE p.id IN (SELECT patient FROM asthmaCohort)

```

**Race Information**

-   Patient's race distribution with average age are described in the following query.

```{sql connection=con}
SELECT p.race, p.gender AS gender, COUNT(*) AS number_of_patient,  ((COUNT(*)*100)/(SELECT COUNT(*) FROM asthmaCohort)) AS percentage 
FROM patients AS p
INNER JOIN 
asthmaCohort AS a
ON a.patient=p.id
WHERE p.deathdate IS NULL
GROUP BY p.race, ROLLUP(p.gender)
ORDER BY 
 (SELECT COUNT(*) FROM patients WHERE race=p.race) DESC,
 p.race,
 CASE WHEN p.gender='M' THEN 0
 WHEN p.gender='F' THEN 1
 ELSE 2
 END


```

**Smoking Status**

-   The patient distribution related to smoking status is described as below.
    -   The smoking ratio among asthma cohort patient group is 87%.

```{sql connection=con}
SELECT (COUNT(DISTINCT (a.patient))*100)/(SELECT COUNT(*) FROM asthmaCohort) AS smoking_ratio
FROM asthmaCohort AS a
INNER JOIN
observations AS o
ON a.patient=o.patient
INNER JOIN
careplans AS c
ON a.patient=c.patient
WHERE o.code='72166-2' AND (c.stop IS NULL OR c.stop >='2022-01-01')
```

**Insurance Status**

-   The patient distribution related to insurance status is described as below. (no-insurance id = e03e23c9-4df1-3eb6-a62d-f70f02301496)
    -   Approximately 11% patients were not in any insurance plans in year 2022.

```{sql connection=con}
SELECT p.name AS insurance_name, COUNT(a.patient) AS number_of_patients, (COUNT(a.patient)*100)/(SELECT COUNT(*) FROM asthmaCohort) AS ratio
FROM asthmaCohort AS a
INNER JOIN
payer_transitions AS pt
ON a.patient=pt.patient
INNER JOIN 
payers AS p
ON p.id=pt.payer
WHERE pt.end_date>='2022-01-01'
GROUP BY p.name
ORDER BY (COUNT(a.patient)*100)/(SELECT COUNT(*) FROM asthmaCohort) DESC

```

#### 3) Asthma Patient Treatment & Care Status Analysis

-   To evaluate how asthma patients were treated and/or managed since year 2022, the following analysis was conducted.
    -   From the careplans table/description column: containing 'Asthma' or 'asthma'. --\> to see how many asthma patients have active careplans.
    -   From the immunizations table/description column: containing 'Pneumococcal' or 'PCV'. --\> to see how many asthma patients have been immunized with pneumococcal vaccine.
    -   From the medications table:\
        --\> to see how many asthma patients were prescribed with asthma medications.

##### 3-1) Active Careplan Status

-   According to the following query, a total 1,381 patients have active careplan, which is 60% of the total asthma patient cohort.

```{sql connection=con}
SELECT COUNT(DISTINCT c.patient) AS patients_with_careplan, (COUNT(DISTINCT c.patient)*100)/(SELECT COUNT(*) FROM asthmaCohort) AS ratio

FROM careplans AS c
INNER JOIN 
asthmaCohort AS a
ON a.patient=c.patient
WHERE (c.description LIKE '%asthma%' OR c.description LIKE '%Asthma%') 
AND (c.stop IS NULL OR c.stop >='2022-01-01')
```

##### 3-2) Immunization Status

-   488 patients from the asthma cohort was immunized with pneumococcal vaccine, which is 21% of the total cohort group.
-   In particular, 57 patients from this cohort were vaccinated since the year 2021, which represents approximately 2% of this cohort.

```{sql connection=con}
WITH pcv_vaccine AS (
SELECT i.patient, i.date
FROM immunizations AS i
INNER JOIN
asthmaCohort AS a
ON a.patient=i.patient
WHERE (i.description LIKE '%PCV%' OR i.description LIKE '%Pneumococcal%')
)
SELECT COUNT(DISTINCT patient) AS immunized_patient, (COUNT(DISTINCT patient)*100)/(SELECT COUNT(*) FROM asthmaCohort) AS immunized_ratio, (SELECT(COUNT(DISTINCT pcv_vaccine.patient)) FROM pcv_vaccine WHERE date >='2021-01-01') AS immunized_since_2021, ((SELECT(COUNT(DISTINCT pcv_vaccine.patient)) FROM pcv_vaccine WHERE date >='2021-01-01')*100)/(SELECT COUNT(*) FROM asthmaCohort) AS immunized_since2021_ratio
FROM pcv_vaccine

```

##### 3-3) Medication Status

-   A total of 1591 patients had medications prescribed related to asthma control (rescue inhaler & maintenance inhaler), representing 69% of the total cohort.
-   All patients who had maintenance inhaler were also prescribed with rescue inhalers. (The number of patients with maintenance inhaler without the rescue inhaler was 0.)
-   Patients who had prescription with the stop date after 2022-01-1 were included in this description.

```{sql connection=con}
WITH maintenance_med AS (
SELECT m.code, m.description, m.patient
FROM medications AS m
INNER JOIN asthmaCohort AS a
ON a.patient=m.patient
WHERE code IN (95996, 349094, 616830, 1998774, 896209, 896006, 966529, 746815, 966524, 1547660, 1998772, 1536144) AND stop >='2022-01-01'
), 
rescue_med AS (
SELECT r.code, r.description, r.patient 
FROM medications AS r
INNER JOIN asthmaCohort AS a
ON a.patient=r.patient
WHERE code IN (630208, 859088, 745752, 351137, 351136, 245314, 2123111, 746763) AND stop >='2022-01-01'
)

SELECT 'maintenance_medication' AS medication_type, COUNT(DISTINCT m.patient) AS patient_count, (COUNT(DISTINCT m.patient)*100)/(SELECT COUNT(*) FROM asthmaCohort) AS ratio
FROM maintenance_med AS m

UNION

SELECT 'rescue_medication' AS medication_type, COUNT(DISTINCT r.patient) AS patient_count, (COUNT(DISTINCT r.patient)*100)/(SELECT COUNT(*) FROM asthmaCohort) AS ratio
FROM rescue_med AS r

UNION 

SELECT 'both_medication' AS medication_type, COUNT(DISTINCT m.patient) AS patient_count, (COUNT(DISTINCT m.patient)*100)/(SELECT COUNT(*) FROM asthmaCohort) AS ratio
FROM maintenance_med AS m
WHERE m.patient IN(
SELECT DISTINCT r.patient
FROM rescue_med AS r
)

```

-   From the maintenance inhaler medication, '60 ACTUAT Fluticasone propionate 0.25 MG/ACTUAT / salmeterol 0.05 MG/ACTUAT Dry Powder Inhaler' was the most prescribed medication.

```{sql connection=con}
WITH maintenance_med AS (
SELECT m.code, m.description, m.patient
FROM medications AS m
INNER JOIN asthmaCohort AS a
ON a.patient=m.patient
WHERE code IN (95996, 349094, 616830, 1998774, 896209, 896006, 966529, 746815, 966524, 1547660, 1998772, 1536144) AND stop >='2022-01-01'
)

SELECT description, COUNT(DISTINCT patient) AS number_of_patient FROM maintenance_med
WHERE patient IN (SELECT * FROM asthmaCohort)
GROUP BY description
ORDER BY COUNT(DISTINCT patient) DESC
```

-   From the rescue inhaler medication, 'albuterol 0.83 MG/ML Inhalation Solution' was the most prescribed medication.

```{sql connection=con}
WITH rescue_med AS (
SELECT r.code, r.description, r.patient 
FROM medications AS r
INNER JOIN asthmaCohort AS a
ON a.patient=r.patient
WHERE code IN (630208, 859088, 745752, 351137, 351136, 245314, 2123111, 746763) AND stop >='2022-01-01'
)

SELECT description, COUNT(DISTINCT patient) AS number_of_patient 
FROM rescue_med
WHERE patient IN (SELECT * FROM asthmaCohort)
GROUP BY description
ORDER BY COUNT(DISTINCT patient) DESC
```

#### 4) Severity of Asthma Condition

-   In order to evaluate the severity of each patients from the asthma cohort, the number of encounters during year 2022 was calculated as below.
-   The higher the number of encounters related to asthma could indicate that the patient is more severe with the asthma condition.

```{sql connection=con}
SELECT e.patient, COUNT(e.description) AS number_of_encounters
FROM encounters AS e
WHERE (e.start BETWEEN '2022-01-01' AND '2022-12-31') AND (e.patient IN (SELECT * FROM asthmaCohort)) AND (reasondescription LIKE '%Asthma%' OR reasondescription LIKE '%asthma%')
GROUP by patient
ORDER BY COUNT(e.description) DESC


```

## IV. Results & Summary

-   Analyses were conducted to study the asthma managment status in year 2022.
-   The patient cohort was defined as patients with history of asthma-related encounters or medications in year 2022.
-   The average age of this patient cohort was 45, and male/female ratio were similar (46% & 52%).
-   Approximately 11% of these patients did not have insurance.
-   60% of them had active careplans regarding asthma management.
-   21% of the patient group were immunized with PCV vaccine at least once in the past, and only 2% of theme were vaccinated since year 2021.
-   69% of patients had medications regarding asthma inhalers.
